Processing genome scale tabular data with wormtable

### 2013-08-17 docs/paper/paper.tex
Spell checked.

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
rows of interest. This allows us to efficiently seek to an 
arbitary location in the table and to read a given number 
of rows sequentially from this point.

@@

Full table scans are often uneccessary in wormtable. For many tasks,
creating an appropriate index allows us to seek directly to the 
rows of interest. To illustrate this, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m40s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

@@

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevent pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, compression is a double-edged sword: 
it does lead to much smaller files, which reduces the number of random seeks 
required.  On the other hand, however, decompression is expensive and must 
be repeated each time a block is accessed unless some sophisticated
caching mechanism is devised.

@@

\section*{Discussion}
The problems of enabling efficient random access to rows and 
avoiding the large overhead of parsing text are well understood, 
and efforts to address them are proceeding in parallel for 
different file formats. 
BCF, for example, is the binary version of the VCF format discussed 
above, in which values within rows are stored in a packed binary 
format.  Similiarly, BigBed and 
BigWig~\cite{kzbhk10} are compressed binary versions of the BED 
and WIG file formats, which offer efficient random access along 
with the ability to operate over a network.

@@

The library supports efficient access to any data stored in 
wormtable format, and currently provides conversion utilitites 
for the VCF and GTF formats.
The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, SAM and BED to wormtable format. 
Such tools are not difficult to develop, since all that is required is a parser
for the format in question written in Python.

>>>>

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
rows of interest. This allows us to efficiently seek to an 
arbitrary location in the table and to read a given number 
of rows sequentially from this point.

@@

Full table scans are often unnecessary in wormtable. For many tasks,
creating an appropriate index allows us to seek directly to the 
rows of interest. To illustrate this, we repeated the example 
of counting transitions and transversions using an index 
on the REF and ALT columns.
The index required 3m40s to build and 
consumed 66MB of storage space. Counting the number of 
transitions and transversions using this index required 
less than a second.

@@

This difference is well illustrated by immediately repeating 
the same experiment, so that all the relevant pages are 
in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
2 seconds for wormtable. Thus, compression is a double-edged sword: 
it does lead to much smaller files, which reduces the number of random seeks 
required.  On the other hand, however, decompression is expensive and must 
be repeated each time a block is accessed unless some sophisticated
caching mechanism is devised.

@@

\section*{Discussion}
The problems of enabling efficient random access to rows and 
avoiding the large overhead of parsing text are well understood, 
and efforts to address them are proceeding in parallel for 
different file formats. 
BCF, for example, is the binary version of the VCF format discussed 
above, in which values within rows are stored in a packed binary 
format.  Similarly, BigBed and 
BigWig~\cite{kzbhk10} are compressed binary versions of the BED 
and WIG file formats, which offer efficient random access along 
with the ability to operate over a network.

@@

The library supports efficient access to any data stored in 
wormtable format, and currently provides conversion utilities 
for the VCF and GTF formats.
The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, SAM and BED to wormtable format. 
Such tools are not difficult to develop, since all that is required is a parser
for the format in question written in Python.
